Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-06-17 paper.tex
Updates from Overleaf

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$. %\rljf{How?}
Similar to the decoder $\psi$, we use the vector representations $f_t \in \mathbb{R}^k$ computed by the LSTM encoder after processing $X_{-M\dots t}$ for $t = -M, \dots, -1$, and then compute predictions using the softmax rule
\begin{equation}
	P_\eta(X_t = s_i|X_{1\dots t-1}, Z) \propto \exp((W_{o'} f_t)_i)
\end{equation}
where $W_{o'} \in \mathbb{R}^{|S|\times k}$ is another parameter matrix.

@@

During optimization, we add the cross-entropy term $- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$ to the PRD objective~(\ref{eq:bound}) \rljf{Why?}, so that the full training objective comes out to:
\begin{equation}\label{eq:bound-augmented}
	\min_{\phi, \psi, q,\eta} \left[\mathbb{E}_{\finitepast, \finitefuture} \left[ -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} \left[ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] - \log P_\eta(\finitepast) \right]\right].
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
Again by Gibbs' inequality and Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}, this is minimized when $P_\eta$ represents the true distribution over length-$M$ blocks $P(\finitepast)$, $P_\phi(Z|\finitepast)$ describes an optimal code for the given $\lambda$, $q$ is its marginal distribution, and $P_\psi(\finitefuture|Z)$ is the Bayes-optimal decoder.
For approximate solutions to this augmented objective, the inequalities~(\ref{eq:bound1-new}-\ref{eq:bound2-new}) will also remain true due to Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}.

@@

Training used Early Stopping using the development set, as described in Section~\ref{sec:parameter-estimation}.
Models were trained on a TITAN Xp graphics card.
\mhahn{TODO number of epochs, training time}

@@

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{code/figures/even-info-varyM.pdf}

%\includegraphics[width=0.45\textwidth]{code/figures/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process (left) and the Random Insertion Process (right), varying $M=5$ (blue), $10$ (red), $15$ (green). Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Compare Figure~\ref{fig:even} for results from full NPRD.}\label{fig:samples-ptb}
\end{figure*}

>>>>

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$. 
Similar to the decoder $\psi$, we use the vector representations $f_t \in \mathbb{R}^k$ computed by the LSTM encoder after processing $X_{-M\dots t}$ for $t = -M, \dots, -1$, and then compute predictions using the softmax rule
\begin{equation}
	P_\eta(X_t = s_i|X_{1\dots t-1}, Z) \propto \exp((W_{o'} f_t)_i)
\end{equation}
where $W_{o'} \in \mathbb{R}^{|S|\times k}$ is another parameter matrix.

@@

In order to optimize parameters for estimation of $P_\eta$, we add the cross-entropy term $- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$ to the PRD objective~(\ref{eq:bound}) during optimization, %\rljf{Why?},
so that the full training objective comes out to:
\begin{equation}\label{eq:bound-augmented}
	\min_{\phi, \psi, q,\eta} \left[\mathbb{E}_{\finitepast, \finitefuture} \left[ -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} \left[ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] - \log P_\eta(\finitepast) \right]\right].
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
Again by Gibbs' inequality and Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}, this is minimized when $P_\eta$ represents the true distribution over length-$M$ blocks $P(\finitepast)$, $P_\phi(Z|\finitepast)$ describes an optimal code for the given $\lambda$, $q$ is its marginal distribution, and $P_\psi(\finitefuture|Z)$ is the Bayes-optimal decoder.
For approximate solutions to this augmented objective, the inequalities~(\ref{eq:bound1-new}-\ref{eq:bound2-new}) will also remain true due to Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}.

@@

Training used Early Stopping using the development set, as described in Section~\ref{sec:parameter-estimation}.
Models were trained on a TITAN Xp graphics card.
\mhahn{TODO number of epochs, training time comparable to OCF}

On English word-level modeling, training took a median number of 9 epochs (max 467 epochs) and 5 minutes (max 126 minutes).

@@

In Figure~\ref{fig:samples-ptb}, we provide four sample outputs from English word-level modeling with three values of $\log \frac{1}{\lambda}$ ($1.0$, $3.0$, $5.0$), corresponding to low ($1.0$), medium ($3.0$), and high ($5.0$) rates  (compare Figure~\ref{fig:wordlevel-fit-1}).
We obtained sample sequences by selecting the first 32 sequences $\finitepast\finitefuture$ (at $M=15$) from the Penn Treebank validation set, and selected four examples where the variation in cross-entropy values at $X_0$ was largest between the three models.

Across these samples, models generated at $\log \frac{1}{\lambda} = 5$ show lower cross-entropy on the first future observation $X_0$, as these codes have higher rates.
For instance, in the first sample, the cross-entropy on the first word \emph{Jersey} is lowest for the code with the higher rate; indeed, this word is presumably strongly predicted by the preceding sequence \emph{...Sen. Billd Bradley of New}.
Codes with higher rates are better at encoding such predictive information from the past.

\begin{figure*}
\centering
\textbf{Past:} within minutes after the stock market closed friday , i called sen. bill bradley of new
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_0.png}

\textbf{Past:} asset you own would be worth more if the government suddenly announced that if you sold
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_20.png}

\textbf{Past:} he thinks , when the white house is forced to ask for higher taxes to meet
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_31.png}

\textbf{Past:} to be dealt with by the oct. 15 deadline , and these senate democrats refused to
\includegraphics[width=0.9\textwidth]{code/nprd/figures/sample_6.png}

	\caption{Four example outputs from English word-level modeling, with low rate ($\log \frac{1}{\lambda} = 1$; red, dotted), medium rate ($\log \frac{1}{\lambda} = 3$; green, dashed), high rate ($\log \frac{1}{\lambda} = 5$; blue, solid). For each sample, we provide the prior context $\finitepast$ (top), and the per-word cross-entropies on the future words $\finitefuture$ (bottom).
	}\label{fig:samples-ptb}
\end{figure*}
